---
title: "SoilHydro: Van Genuchten Water Retention Curve Tools"
subtitle: "Fitting, Diagnostics, Hydrophysical Metrics, and Visualization"
author: "Erick Gutierrez / Auburn University"
output:
  html_vignette:
    toc: true
    number_sections: true
  github_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{SoilHydro: Van Genuchten Water Retention Curve Tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
```

# Overview

**SoilHydro** is an R package for fitting **Van Genuchten** water retention curves (WRC) and computing practical hydrophysical summaries used in soil physics and agronomy. The package:

- standardizes suction units (accepts **kPa** or **hPa**, fits internally in kPa);
- fits the Van Genuchten model per treatment/group with bounded L‑BFGS‑B optimization;
- predicts water content \( \theta(h) \) on arbitrary suction grids;
- derives **water points** (e.g., Field Capacity at 10 kPa, Permanent Wilting Point at 1500 kPa, **Available Water Content**);
- partitions porosity into **macro/meso/micro** classes from capillarity;
- and provides clean **ggplot2** visualizations for curves and pore-size bars.

We assume familiarity with basic soil physics concepts (\( \theta_r, \theta_s, \alpha, n, m \)) and laboratory WRC data.

---

## Live web app

You can try **all SoilHydro functions interactively in your browser**—no installation needed:

👉 [SoilHydro Shiny web app](https://3n0c6z-erick-gutierrez.shinyapps.io/SoilHydro/)

The app mirrors the package workflow (fitting, prediction, water points, pore-size classes, and visualization).

---

```{r pkg-versions, include=FALSE}
sessionInfo()
```

# Installation

```r
# Install from GitHub using devtools
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}

devtools::install_github("egubens/SoilHydro")
```

**Requires**: R ≥ 3.6; Depends on base **stats** and **ggplot2**.

# Data requirements

Your long/tidy dataset should include at least:

- an **ID/treatment** column (optional for single‑profile fits), e.g. `Product`;
- a **theta** column with volumetric water content \( \theta \) (m^3 m^{-3});
- a **suction** column `h` (either **kPa** or **hPa**).

Pass the units of your raw suction column via `units = "kPa"` or `"hPa"`.

# The Van Genuchten model

We model water content \( \theta(h) \) (with suction \( h \) in **kPa**) as


\[
theta(h) \;=\; \theta_r + \frac{\theta_s - \theta_r}{\bigl(1 + (\alpha h)^n\bigr)^m},
\quad m = 1 - \frac{1}{n}, \; n>1, \; \alpha>0.
\]

The package uses **Mualem's** linkage \( m = 1 - 1/n \) and fits parameters by minimizing the sum of squared errors (SSE) with **L‑BFGS‑B** bounds to enforce \( \theta_s>\theta_r\), \(n>1\), \( \alpha>0\).

# Quick start with simulated data

Below we simulate three treatments with “true” parameters, sample common laboratory suctions in **hPa**, add small noise, and run the full workflow.

```{r quickstart, message=FALSE, warning=FALSE}
library(SoilHydro)
library(ggplot2)

set.seed(42)

# 1) "True" parameters (kPa units)
true_params <- list(
  Control     = list(theta_r = 0.05, theta_s = 0.42, alpha = 0.030, n = 1.42),
  ArtiMicro05 = list(theta_r = 0.02, theta_s = 0.44, alpha = 0.100, n = 1.55),
  Persist10   = list(theta_r = 0.04, theta_s = 0.46, alpha = 0.060, n = 1.50)
)

# 2) Suction points (hPa)
hPa_grid <- c(1, 10, 30, 50, 75, 100, 220, 340, 440, 650, 1000, 1500, 3000, 8000, 15000)

# 3) Generate theta(h) + noise
sim_rows <- lapply(names(true_params), function(id){
  p  <- true_params[[id]]
  h_kPa <- hPa_grid / 10
  theta <- SoilHydro::vg_fun_kPa(h_kPa, p$theta_r, p$theta_s, p$alpha, p$n)
  theta_noisy <- pmin(p$theta_s, pmax(0, theta + rnorm(length(theta), sd = 0.006)))
  data.frame(Product = id, hPa = hPa_grid, theta = theta_noisy, stringsAsFactors = FALSE)
})
df <- do.call(rbind, sim_rows)

head(df, 8)
```

## Fit per treatment

```{r fit}
fits <- vg_fit_optim(
  data  = df,
  id    = "Product",
  theta = "theta",
  h     = "hPa",
  units = "hPa"
)
fits
```

## Plot WRC

```{r plot-wrc, fig.height=4.8}
plots <- plot_vg_fits(
  data      = df,
  params_df = fits,
  id        = "Product",
  theta     = "theta",
  h         = "hPa",
  units     = "hPa",
  log_x     = TRUE
)
plots[["Control"]]
```

## Derived metrics: water points and AWC

We define **FC** at 10 kPa and **PWP** at 1500 kPa. **AWC** is \( \theta_{fc} - \theta_{pwp} \).

```{r water-points}
wp <- vg_water_points(fits, id_col = "Product", fc_kPa = 10, pwp_kPa = 1500)
wp
```

## Pore-size distribution (PSD)

By capillarity, the characteristic suction \( \psi \) (kPa) associated with a pore of diameter \( d \) (µm) is approximated by


\[
\psi \;\approx\; \frac{4 \gamma \cos\theta_c}{d} \times 10^{-3},
\]

with surface tension \( \gamma \) (N m^{-1}) and contact angle \( \theta_c \). SoilHydro defaults to \( \gamma \approx 0.0728 \) at 20 °C and \( \cos\theta_c = 1 \). Using cutoffs \( d_{macro}=1000\,\mu\text{m} \) and \( d_{micro}=10\,\mu\text{m} \), the package partitions porosity into:

- **Macro**: \( d>1000\,\mu\text{m} \) (large, drainage/airflow);
- **Meso**: \( 10\text{–}1000\,\mu\text{m} \) (plant‑available water);
- **Micro**: \( d<10\,\mu\text{m} \) (tightly held water).

By default, **residual water** \( \theta_r \) is included within “micro” so that macro+meso+micro \(\approx \theta_s\).

```{r psd-tbl}
psd <- vg_pore_classes(
  fits,
  id_col = "Product",
  percent_basis = "theta_s",
  include_residual_in_micro = TRUE
)
psd
```

### PSD plots

```{r psd-plots, fig.height=4.8}
g_pct <- plot_pore_classes_percent(psd, id_col = "Product")
g_vol <- plot_pore_classes_volume(psd,  id_col = "Product")
g_pct
g_vol
```

# Detailed workflow

## 1. Units and preprocessing

- Specify `units = "kPa"` or `"hPa"` **matching your input column**.
- Internally, SoilHydro converts to **kPa** for fitting; parameters are returned with \( \alpha \) in kPa^{-1}.
- Remove zero/negative suctions when plotting on log axes (the plotting helper already does this).

## 2. Fitting and diagnostics

`vg_fit_optim()` performs bounded least‑squares with sensible starting values per group. Returned diagnostics include **R²** and **RMSE** computed on the fitting data.

**Tips for robust fitting**

- Make sure your dataset spans low to high suctions to identify both \( \theta_s \) and \( \theta_r \).
- If a group fails to converge (`.fit_ok = FALSE`), check units, outliers, and monotonicity of \( \theta(h) \).
- Consider fixing a minimal positive suction (e.g., replace zeros by \(10^{-6}\) kPa) for stability on log scales.

## 3. Prediction grids

Use `vg_predict()` to predict \( \theta \) on custom suction grids (kPa or hPa). This is useful for report tables and smooth plotting with fixed ranges.

```{r predict}
# Predict on a custom kPa grid
grid_kPa <- c(seq(0, 50, by=1), seq(55, 100, by=5), seq(110, 1600, by=10))
pred <- vg_predict(fits, id_col = "Product", new_h = grid_kPa, units = "kPa")
head(pred, 10)
```

## 4. Interpreting water points and AWC

FC at 10 kPa. PWP at 1500 kPa is conventional for many crops. **AWC** helps compare treatments but should be complemented by conductivity/retention hysteresis where available.

## 5. Pore-size classes

The PSD aggregation condenses the \( \theta(h) \) curve into interpretable buckets. You can change the cutoffs via `d_micro_um` and `d_macro_um`, choose percentage basis (`"theta_s"` vs `"available" = theta_s - theta_r`), and decide whether to include \( \theta_r \) in micro.

```{r psd-options}
psd2 <- vg_pore_classes(
  fits, id_col = "Product",
  d_micro_um = 5, d_macro_um = 500,
  percent_basis = "available",
  include_residual_in_micro = FALSE,
  return_residual = TRUE
)
head(psd2)
```

# Visualization options

- `plot_vg_fits(..., log_x = TRUE)` draws one **ggplot** per ID with observed points and a fitted curve.
- `plot_pore_classes_percent()` and `plot_pore_classes_volume()` produce stacked bars with optional horizontal layout (`horiz = TRUE`).

```{r viz-example, fig.height=4.8}
# Show two IDs side-by-side using patchwork if desired (not required by SoilHydro)
# install.packages("patchwork")  # if needed
# library(patchwork)
# plots[["Control"]] + plots[["Persist10"]]
plots[[2]]
``` 

# Shiny app: SoilHydro_App()

The package includes a **Shiny app** that lets users:

- upload CSV/Excel data and map columns;
- fit per‑ID VG parameters;
- view tables for fits, water points, and pore classes;
- plot WRC with observed vs fitted curves;
- visualize PSD as percent or volume bars; and
- download CSVs of fits and predictions.

To run the app, ensure your SoilHydro functions are installed/available, then source and launch the app:

```r
SoilHydro_App()
```

# Troubleshooting

- **No fits converged**: verify units, remove non‑finite values, ensure \( \theta \in [0,1.2] \) (or realistic bounds).
- **R² is NaN**: occurs when all \( \theta \) values are identical; broaden suction range or re‑measure.
- **Unrealistic \( \alpha, n \)**: check data coverage; tighten bounds via `lower`/`upper` in `vg_fit_optim()`.

# API quick reference

- **Fitting/prediction**
  - `vg_fit_optim(data, id, theta, h, units)` → per‑group parameters + diagnostics.
  - `vg_predict(params_df, id_col, new_h, units, filter_id)` → \( \theta(h) \) table.

- **Derived metrics**
  - `vg_water_points(params_df, id_col, fc_kPa=10, pwp_kPa=1500)` → \( \theta_{fc}, \theta_{pwp}, \text{AWC} \).
  - `vg_pore_classes(params_df, id_col, d_micro_um=10, d_macro_um=1000, percent_basis=c("theta_s","available"), include_residual_in_micro=TRUE, return_residual=FALSE)` → macro/meso/micro volumes & %.

- **Plotting**
  - `plot_vg_fits(data, params_df, id, theta, h, units, log_x=TRUE)` → WRC plots by ID.
  - `plot_pore_classes_percent(psd_df, id_col, horiz=FALSE)` → stacked % bars.
  - `plot_pore_classes_volume(psd_df, id_col, horiz=FALSE)` → stacked volume bars.

# Reproducibility

For manuscripts, record R and package versions and set a seed for any simulation steps (as done above). Export figures with explicit sizes (e.g., `ggsave(..., width=8, height=5, units='cm', dpi=300)`) when preparing journal‑ready graphics.

# References

- Van Genuchten, M. Th. (1980). A closed-form equation for predicting the hydraulic conductivity of unsaturated soils. *Soil Science Society of America Journal*, 44, 892–898.
- Mualem, Y. (1976). A new model for predicting the hydraulic conductivity of unsaturated porous media. *Water Resources Research*, 12, 513–522.
